---
title: "Titanic_Survival"
author: "Ivan"
date: "24 août 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r librabry, echo = FALSE}
library(data.table)
library(tidyr)
library(reshape2)
library(ggplot2)
library(class)
library(mice)
library(VIM)
```
## Get familiar with the dataset

Of course we want to read the data and put it in a structure so we can
start our analysis. Here we chose a data.table, which is, to keep it simple, an
enhanced data.frame. If you are interested, I'm sure you will succeed in finding
more on your own (Google).

```{r init, cache = T}
trainDt <- fread("train.csv", sep = ",", header = T,
                 na.strings = "", stringsAsFactors = T)
testDt <- fread("test.csv", sep = ",", header = T,
                na.strings = "", stringsAsFactors = T)
wholeDt <- rbind(trainDt, testDt, fill = T)
```

Now that we have all our data in one data.table, let's take a peek at it
```{r printData}
wholeDt
summary(wholeDt)
```

What are the remarkable features ? First, we see that some variables have missing
values. This is going to be quite a problem so we will deal with it later.
Second, some variables, namely Name, Ticket and Cabin, seems to contain more
than one information. Our first task will be to transform the data so we can use
this information.

## Transform the data
For this we will use the tidyr package. Let's get started.
### Name
```{r transformName}
wholeDt <- separate(wholeDt,
                     Name,
                     c('Family', 'Title'),
                     ", ")
wholeDt[, `:=` (Title = as.factor(gsub("\\..*", "", Title)),
                Family = as.factor(Family))]
head(wholeDt, 2)
```

We splitted the Name column into the Family and Title column. In the process we
discarded passengers first names as we can't see how being named Francis would 
ber relevant to our survival on the Titanic. We might be wrong though... Now one
can notice that we might have duplicate on those two column, like a father and
his son but the remaining variables hold plenty of information to distinct them.

This leaves us with two factor variables that might be related to survival. The
Family one is tricky so let's start with the Title one.
```{r peekTitle}
table(wholeDt$Title)
```
Those factors are not satisfying as some are pretty much the same and some are
too rare to be of any use. So we are going to regroup Don, Dona, Jonkheer, Lady,
Sir and Countess into a Noble factor, Col, Capt, Dr, Major and Rev into
Honorific and finally reassign Ms and Mlle to Miss and Mme to Mrs.
```{r changeTitle}
nobles <- c("Don", "Dona", "Jonkheer", "Lady", "Sir","the Countess")
honorifics <- c("Col", "Capt", "Dr", "Major", "Rev")
intoMiss <- c("Ms", "Mlle")
intoMrs <- c("Mme")
wholeDt[Title %in% intoMiss , Title := "Miss"]
wholeDt[Title %in% intoMrs, Title := "Mrs"]
wholeDt[Title %in% nobles, Title := "Noble"]
wholeDt[Title %in% honorifics, Title := "Honorific"]
wholeDt[, Title := droplevels(Title)]
```
Now we are good with this Title variable. We still have a tricky problem with
our Family variable. Can you guess it ?

What if two family have the same last name ? There won't be a reliable way to
distinct them. There are also cases where different member of the same family
don't share the same family name, a maried sister for example. 
Let's create a family count and hopefully both variable will contain enough
information. Their is room for amelioration here, for example by getting the
maried women maiden name.
```{r createFamilyCount, warnings = FALSE, message = FALSE}
wholeDt[, FamilyCount := SibSp + Parch + 1]
```

### Cabin
This one is easier we just want to split the Cabin letter and number as it is
probably related to a position on the boat. Some Cabin values contain more than
one Cabin, we'll just take the first one as the other are closely related.

```{r changeCabin}
wholeDt[, Cabin := sapply(strsplit(as.character(Cabin),
                                  " "),
                         function(x) (x[1]))]
wholeDt <- separate(wholeDt, Cabin, c("CabinLetter", "CabinNumber"), 1)
wholeDt[CabinNumber == "", CabinNumber := NA]
wholeDt[,`:=` (CabinLetter = as.factor(CabinLetter),
               CabinNumber = as.integer(CabinNumber))]
```

### Ticket
From this variable it might be interesting to know how many people share the
same ticket. To do this we will create a new column TicketShare which will
contain this total number of people.
```{r TicketShare}
wholeDt[, TicketShare := .N, by = Ticket]
```

## Missing value
Let's find out which variables have missing value and how much are missing. 
```{r NAcols}
colSums(is.na(wholeDt))
```

Our 418 Survived missing value are perfectly normal as we are trying to guess
those. 
### Fare
Passenger 1044 Fare is missing, we will try to input it based on other variables
value. It is probably safe to assume Fare is related to Pclass and maybe to
Embarked. Let's confirm that with a heatmap.

```{r fareCor}
fareCorMat <- as.data.table(
  melt(
    cor(sapply(
      wholeDt[complete.cases(Age,
                             Fare,
                             Embarked) & Age > 30,
              -c("Survived","CabinLetter", "CabinNumber"),
              with = FALSE],
      as.numeric))))
ggplot(fareCorMat[], aes(x=Var1, y=Var2, fill=value)) + 
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab",
   name="Pearson\nCorrelation") +
  theme_minimal()+ 
  geom_tile(color = "white") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Since Cabin and Survived information are missing for this passenger we excluded
it from the heatmap. We can see that as we supposed Pclass and Embarked do
influence Fare but Sex, Age and having a Family also does. Let's look at the
Fare distibution for a related population
```{r fareDistribution}
ggplot(wholeDt[Sex == "male" &
                 Pclass == 3 &
                 Embarked == "S" &
                 FamilyCount == 1 &
                 TicketShare == 1 &
                 !is.na(Fare)],
       aes(Fare)) +
  geom_density(fill = "green", alpha=0.5) +
  geom_vline(
    aes(xintercept = median(Fare, na.rm=T)),
    colour='red',
    linetype='dashed')
wholeDt[PassengerId == 1044, Fare := wholeDt[Sex == "male" &
                                               Pclass == 3 &
                                               Embarked == "S" &
                                               FamilyCount == 1,
                                             median(Fare, na.rm=T)]]
```

Judging from this distribution it is safe to input the median value computed as
Fare value for our Passenger 1044. As you can see we left out the age variable
even though it is impacting, but we don't have enough data for eldery people in
this group.

### Embarked
```{r checkEmbarkedNA}
wholeDt[is.na(Embarked)]
```
These two have the same ticket number so it is likely that they boarded in the 
same place. Let's look at tickets to see if tickets alike all boarded in the 
same place also.

```{r checkSimilarTicket}
wholeDt[Ticket %like% 1135]
```
So it seems our Passenger boarded either in C or in S. Let's get deeper into
analyse like we did for Fare. But in this case all other data are filled.

```{r ticketCor}
ticketCorMat <- as.data.table(
  melt(
    cor(sapply(
      wholeDt[complete.cases(Age,
                             Ticket,
                             Embarked),
              -c("Survived","CabinLetter", "CabinNumber"),
              with = FALSE],
      as.numeric))))
ggplot(ticketCorMat[], aes(x=Var1, y=Var2, fill=value)) + 
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab",
   name="Pearson\nCorrelation") +
  theme_minimal()+ 
  geom_tile(color = "white") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
As we could expect, Embarked depends mainly on Pclass and Fare and more lightly
on Sex. Let's look at related Passengers in those terms then.

```{r relatedEmbarked}
ggplot(wholeDt[Sex == "female" &
                 Pclass == 1 &
                 TicketShare == 2 &
                 !is.na(Embarked)],
       aes(Embarked, Fare)) +
  geom_boxplot()
wholeDt[is.na(Embarked), Embarked := C]
```
Looking at this boxplot, we can input the Embarked value with C as it fits best.
Once again we are left with the most difficult, Age and heavily missing Cabin
features. Let's user mice for that.

### Age and Cabin with mice

testLevels <- sapply(trainDt[, list(Sex, CabinLetter, Embarked)], levels)
trainDt[, `:=` (Sex = as.numeric(Sex),
               CabinLetter = as.numeric(CabinLetter),
               Embarked = as.numeric(Embarked))]
trainDt[, `:=` (CabinLetter = NULL, CabinNumber = NULL)]
corMat <- melt(cor(trainDt[complete.cases(Age, Fare, Embarked)]))
ggplot(corMat, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile()


```{r functions, cache = T}
normalize <- function(x) {
    return ((x - min(x)) / (max(x) - min(x)))
  }
```


```{r learning}
#trainDt <- as.data.table(sapply(trainDt[complete.cases(Age, Fare, Embarked)], normalize))
#testDt <- as.data.table(sapply(testDt, normalize))
#predSurv <- knn(train = trainDt, test = testDt, cl = trainDt[, Survived], k = round(sqrt(nrow(trainDt))))
```